Containers and R---the Rockerverse and beyond

### 2020-01-14 manuscript.Rmd
Docker = lingua franca, also add labels, just in case, small fixes

# Introduction

The R community keeps growing. This can be seen in the number of new packages on CRAN, which keeps on growing exponentially \citep{cran:2019}, but also in the numbers of conferences, open educational resources, meetups, unconferences, and companies taking up, as exemplified by the useR! conference series\footnote{\href{https://www.r-project.org/conferences/}{https://www.r-project.org/conferences/}}, the global growth of the R and R-Ladies user groups\footnote{\href{https://www.r-consortium.org/blog/2019/09/09/r-community-explorer-r-user-groups}{https://www.r-consortium.org/blog/2019/09/09/r-community-explorer-r-user-groups}, \href{https://www.r-consortium.org/blog/2019/08/12/r-community-explorer}{https://www.r-consortium.org/blog/2019/08/12/r-community-explorer}}, or the foundation and impact of the R&nbsp;Consortium\footnote{\href{https://www.r-consortium.org/news/announcements}{https://www.r-consortium.org/news/announcements}, \href{https://www.r-consortium.org/blog/2019/11/14/data-driven-tracking-and-discovery-of-r-consortium-activities}{https://www.r-consortium.org/blog/2019/11/14/data-driven-tracking-and-discovery-of-r-consortium-activities}}.
All this cements the role of R as the _lingua&nbsp;franca_ of statistics, data visualisation, and computational research.
Coinciding with the rise of R was the advent of [Docker](https://en.wikipedia.org/wiki/Docker_(software)) as a general tool for distribution and deployment of server applications.
Combining both these topics, the _Rocker&nbsp;Project_ ([https://www.rocker-project.org/](https://www.rocker-project.org/)) provides images with R (see the next Section for more details).
The considerable uptake and continued evolution of the Rocker Project has lead to numerous projects extending or building upon Rocker images, ranging from reproducible research to production deployments.
This article presents this _Rockerverse_ of projects across all development stages: early demonstrations, working prototypes, and mature products.
We also introduce related activities connecting the R language and environment with other containerisation solutions.
The main contributions is a coherent picture of the current lay of the land of using containers in, with, and for R.

@@

# Containerization and Rocker

@@

The key accomplishment of Docker as an "application" is to make a "bundled" aggregation of software (the so-called "image") available to any system equipped to run Docker, without requiring much else from the host besides the actual Docker application installation.
This is a rather attractive proposition and Docker's very easy to use user interface has lead to widespread adoption and use of Docker in a variety of domains, e.g., cloud computing infrastructure \citep{Bernstein2014}, data science \citep{boettiger_introduction_2015}, and edge computing \citep{alam_orchestration_2018} [more examples!].
It provided to be a natural match for "cloud deployment" which runs, or at least appears to run, "seamlessly" without much explicit reference to the underlying machine, architecture or operating system: containers are portable and can be deployed with very little in terms of dependencies on the host system---only the container runtime is required.
Images are normally build from plain text documents called `Dockerfile`.
A `Dockerfile` has a specific set of instructions to create and document a well-defined environment, i.e., install specific software and expose specific ports.

For statistical computing and analysis centered around R, the Rocker Project has provided a variety of Docker containers since its start in 2014 \citep{RJ-2017-065}.
The Rocker Project provides several lines of containers spanning to from building blocks with `R-release` or `R-devel`, via containers with [RStudio&nbsp;Server](https://rstudio.com/products/rstudio/) and [Shiny&nbsp;Server](https://rstudio.com/products/shiny/shiny-server/), to domain-specific containers such as [`rocker/geospatial`](https://github.com/rocker-org/geospatial) \citep{rocker_geospatial_2019}.
These containers form _image stacks_, building on top of each other for better maintainability and 
Also of note is a series of "versioned" containers which match the R release they contain with the _then-current_ set of packages via the MRAN Snapshot views of CRAN \citep{microsoft_cran_2019}.
The Rocker Project's impact and importance was acknowledged by the Chan Zuckerberg Initiative's _Essential Open Source Software for Science_, who provide findung for the projects's sustainable maintenance, community growth, and targeting new hardware platforms including GPUs \citep{chan_zuckerberg_initiative_maintaining_2019}.

Docker is not the only containerisation software.
An alternative stemming from the domain of high-perfomance computing is **Singularity** \citep{kurtzer_singularity_2017}.
Singularity can run Docker images, and in the case of Rocker works out of the box if the main process is R, e.g., in `rocker/r-base`, but does not succeed in running images where there is an init script, e.g., in containers that by default run RStudio.
In the latter case, a `Singularity` file, a recipe akin to a `Dockerfile`, needs to be used.
To date, no comparable image stack to Rocker exists on [Singularity Hub](https://singularity-hub.org/).
A further tool for running containers is [**podman**](https://github.com/containers/libpod), which also can build `Dockerfile`s and run Docker images.
Proof of concepts for using podman to build and run Rocker containers exist\footnote{See \href{https://github.com/nuest/rodman}{https://github.com/nuest/rodman} and \href{https://github.com/rocker-org/rocker-versioned/issues/187}{https://github.com/rocker-org/rocker-versioned/issues/187}}.
Yet the prevelance of Docker, especially in the broader user community beyond experts or niche systems, and the vast amount of blog posts and courses for Docker, currently caps specific development efforts for both Singularity and podman in the R community.
This might quickly change when usability and spread increase, or security features such as rootless/unprivileged containers, which both these tools support out of the box, become more sought after.

# Container images

## Images for alternative R distributions

@@

## Bioconductor

@@

## Images for (historic) R versions

@@

## Windows Images

@@

## Non-Debian Linux images

@@

## Data Science images

@@

# Use cases and applications
<!-- from packages closer to Docker to second level packages to more complex use cases -->

@@

## Capture and create environments [@nuest]

@@

## Using R to power enterprise software in production environments

@@

## Deployment to the cloud

@@

## Continuous integration and continuous delivery [@noamross, @ColinFay]

@@

## Common or public work environments

@@

## Using Docker in a commercial Data Science Platform

@@

## Processing [div]

>>>>

# Introduction
\label{intro}

The R community keeps growing. This can be seen in the number of new packages on CRAN, which keeps on growing exponentially \citep{cran:2019}, but also in the numbers of conferences, open educational resources, meetups, unconferences, and companies taking up, as exemplified by the useR! conference series\footnote{\href{https://www.r-project.org/conferences/}{https://www.r-project.org/conferences/}}, the global growth of the R and R-Ladies user groups\footnote{\href{https://www.r-consortium.org/blog/2019/09/09/r-community-explorer-r-user-groups}{https://www.r-consortium.org/blog/2019/09/09/r-community-explorer-r-user-groups}, \href{https://www.r-consortium.org/blog/2019/08/12/r-community-explorer}{https://www.r-consortium.org/blog/2019/08/12/r-community-explorer}}, or the foundation and impact of the R&nbsp;Consortium\footnote{\href{https://www.r-consortium.org/news/announcements}{https://www.r-consortium.org/news/announcements}, \href{https://www.r-consortium.org/blog/2019/11/14/data-driven-tracking-and-discovery-of-r-consortium-activities}{https://www.r-consortium.org/blog/2019/11/14/data-driven-tracking-and-discovery-of-r-consortium-activities}}.
All this cements the role of R as the _lingua&nbsp;franca_ of statistics, data visualisation, and computational research.
Coinciding with the rise of R was the advent of [Docker](https://en.wikipedia.org/wiki/Docker_(software)) as a general tool for distribution and deployment of server applications---in fact, Docker can be called the _lingua&nbsp;franca_ of describing computing environments and packaging software.
Combining both these topics, the _Rocker&nbsp;Project_ ([https://www.rocker-project.org/](https://www.rocker-project.org/)) provides images with R (see the next Section for more details).
The considerable uptake and continued evolution of the Rocker Project has lead to numerous projects extending or building upon Rocker images, ranging from reproducible research to production deployments.
This article presents this _Rockerverse_ of projects across all development stages: early demonstrations, working prototypes, and mature products.
We also introduce related activities connecting the R language and environment with other containerisation solutions.
The main contributions is a coherent picture of the current lay of the land of using containers in, with, and for R.

@@

# Containerization and Rocker
\label{containerisation}
\label{rocker}

@@

The key accomplishment of Docker as an "application" is to make a "bundled" aggregation of software (the so-called "image") available to any system equipped to run Docker, without requiring much else from the host besides the actual Docker application installation.
This is a rather attractive proposition and Docker's very easy to use user interface has lead to widespread adoption and use of Docker in a variety of domains, e.g., cloud computing infrastructure \citep[e.g.,][]{Bernstein2014}, data science \citep[e.g.,][]{boettiger_introduction_2015}, and edge computing \citep[e.g.,][]{alam_orchestration_2018}.
It provided to be a natural match for "cloud deployment" which runs, or at least appears to run, "seamlessly" without much explicit reference to the underlying machine, architecture or operating system: containers are portable and can be deployed with very little in terms of dependencies on the host system---only the container runtime is required.
Images are normally build from plain text documents called `Dockerfile`.
A `Dockerfile` has a specific set of instructions to create and document a well-defined environment, i.e., install specific software and expose specific ports.

For statistical computing and analysis centered around R, the Rocker Project has provided a variety of Docker containers since its start in 2014 \citep{RJ-2017-065}.
The Rocker Project provides several lines of containers spanning to from building blocks with `R-release` or `R-devel`, via containers with [RStudio&nbsp;Server](https://rstudio.com/products/rstudio/) and [Shiny&nbsp;Server](https://rstudio.com/products/shiny/shiny-server/), to domain-specific containers such as [`rocker/geospatial`](https://github.com/rocker-org/geospatial) \citep{rocker_geospatial_2019}.
These containers form _image stacks_, building on top of each other for better maintainability (i.e. smaller `Dockerfile`s), composability, and to reduce build time.
Also of note is a series of "versioned" containers which match the R release they contain with the _then-current_ set of packages via the MRAN Snapshot views of CRAN \citep{microsoft_cran_2019}.
The Rocker Project's impact and importance was acknowledged by the Chan Zuckerberg Initiative's _Essential Open Source Software for Science_, who provide findung for the projects's sustainable maintenance, community growth, and targeting new hardware platforms including GPUs \citep{chan_zuckerberg_initiative_maintaining_2019}.

Docker is not the only containerisation software.
An alternative stemming from the domain of high-perfomance computing is **Singularity** \citep{kurtzer_singularity_2017}.
Singularity can run Docker images, and in the case of Rocker works out of the box if the main process is R, e.g., in `rocker/r-base`, but does not succeed in running images where there is an init script, e.g., in containers that by default run RStudio&nbsp;Server.
In the latter case, a `Singularity` file, a recipe akin to a `Dockerfile`, needs to be used.
To date, no comparable image stack to Rocker exists on [Singularity Hub](https://singularity-hub.org/).
A further tool for running containers is [**podman**](https://github.com/containers/libpod), which also can build `Dockerfile`s and run Docker images.
Proof of concepts for using podman to build and run Rocker containers exist\footnote{See \href{https://github.com/nuest/rodman}{https://github.com/nuest/rodman} and \href{https://github.com/rocker-org/rocker-versioned/issues/187}{https://github.com/rocker-org/rocker-versioned/issues/187}}.
Yet the prevelance of Docker, especially in the broader user community beyond experts or niche systems, and the vast amount of blog posts and courses for Docker, currently caps specific development efforts for both Singularity and podman in the R community.
This might quickly change when usability and spread increase, or security features such as rootless/unprivileged containers, which both these tools support out of the box, become more sought after.

# Container images
\label{images}

## Images for alternative R distributions
\label{alternatives}

@@

## Bioconductor
\label{bioc}

@@

## Images for (historic) R versions
\label{versions}

@@

## Windows Images
\label{windows}

@@

## Non-Debian Linux images
\label{nondebian}

@@

## Data Science images
\label{datascience}

@@

# Use cases and applications
<!-- from packages closer to Docker to second level packages to more complex use cases -->
\label{applications}

@@

## Capture and create environments [@nuest]
\label{envs}

@@

## Using R to power enterprise software in production environments
\label{enterprise}

@@

## Deployment to the cloud
\label{deployment}

@@

## Continuous integration and continuous delivery [@noamross, @ColinFay]
\label{cicd}

@@

## Common or public work environments
\label{workenvs}

@@

## Using Docker in a commercial Data Science Platform
\label{rstudio}

@@

## Processing [div]
\label{processing}
